Real-time modelling of a public transport network to improve arrival-time predictions

### 2017-08-03 realtime-filter/realtime-filter.Rnw

Do this by identifying all intersections,
which can be downloaded from OpenStreetMap \cite{OpenStreetMap_2017}
and segment routes at these points,
as shown in figure~\ref{fig:route_segmentation}.
For now, only using traffic lights and roundabouts as simplification.
Modify \gls{gtfs} data so \emph{shapes} table becomes
\texttt{shapes}, \texttt{segments}, and \texttt{segment\_shapes}.
We assume that if two routes both go through intersections $A$ and $B$,
they follow the same road to do so
(routes retain their own shape, so in cases where this \emph{is not} true,
the shape will still be valid).

Thus as we add new routes, we don't need to create new segments
(except at start/end\ldots see section~\ref{sec:discussion}).

@@

@@

@@

%\begin{xtabular}{lp{6cm}}
\begin{xtabular}{ll}
  \toprule
  &\textbf{Data} \\ \midrule
  $t_k$, $t_c$ & time at observation $k$, $c$ \\
  $\delta_x$ & time since last observation, $t_x - t_{x-1}$\\
  $\bY_k$ & observation $k$ \\
  $\tilde y_k$ & GPS position of vehicle (degrees) \\
  $\phi,\lambda$ & latitude, longitude (radians) \\
  $T^a_j, T^d_j$ & arrival, departure time at stop $j \in \{1, \ldots, J\}$ \\
  $S_n$ & unique identifier of stop $n$ \\
  $s_j$, $s_j^x$, $s_j^p$ & unique identifier of $j$th stop (of relevent route), shape distance traveled, position \\
  $R_m$ & unique identifier of segment $m$  \\
  $r_\ell$, $r_\ell^x$, $r_\ell^p$ & unique identifier of $\ell$th segment (of relevent route), shape distance traveled, path \\
  \bottomrule
  \\ \toprule
  &\textbf{Parameters} \\ \midrule
  $\pi$ & probability of stopping at a stop \\
  $\gamma$ & minimum dwell time (seconds) \\
  $\tau$ & mean dwell time (seconds) \\
  $\rho$ & probability of stopping at an intersection \\
  $\theta_m$ & mean queue time (seconds) at intersection of segment $m$ \\
  $\beta_m$ & mean travel time (seconds) for segment $m$ \\
  $\bbeta$, $\Sigma$ & transit network state (mean, covariance) \\
  $\eta$ & stop size (m) \\
  $\kappa$ & intersection size (m) \\
  $\sigma_x$ & system noise (variability in speed) \\
  $\sigma_y$ & GPS error \\
  $\varepsilon$ & arrival/departure time error \\
  $\xi$ & system noise (Kalman filter) \\
  \bottomrule
  \\ \toprule
  &\textbf{Variables} \\ \midrule
  $\bX_k$ & vehicle state associated with obs $k$ \\
  $x$, $\dot x$ & vehicle distance into trip, velocity \\
  $j$, $\ell$ & stop, segment index \\
  $A_j$, $\tilde D_j$, $D_j$ & arrival, dwell, departure time at stop $j$ \\
  $q_j$, $b_j$ & queue, travel time \\
  $z^{(i)}$ & computed GPS position of particle $(i)$ \\
  $p^{(i)}$ & indicator to whether particle $(i)$ stopped \\
  $w$ & particle weight \\
  $B_m$, $Q_m$ & mean, covariance of travel time of all recent vehicles along segment $m$ \\
  $\mat F_c$ & network state transition matrix \\
  $V$, $v$ & number of vehicles, indexed by $v$ \\
  $I$, $i$ & number of particles (per vehicle), indexed by $i$ \\
  $J$, $j$ & number of stops (conditional on route), indexed by $j$ \\
  $L$, $\ell$ & number of segments (conditional on route), indexed by $\ell$ \\
  $N$, $n$ & total number of stops in network, indexed by $n$ \\
  $M$, $m$ & total number of segments in network, indexed by $m$ \\
  $k$ & index for vehicle observations \\
  $c$ & index for transit network state updates \\
  \bottomrule
  \\ \toprule
  &\textbf{Functions} \\ \midrule
  $f$ & vehicle transition function \\
  $h$ & vehicle measurement function \\
  $g$ & equirectangular projection \\
  \bottomrule
\end{xtabular}

>>>>

Do this by identifying all intersections,
which can be downloaded from \gls{osm} \cite{OpenStreetMap_2017}
and segment routes at these points,
as shown in figure~\ref{fig:route_segmentation}.
For now, only using traffic lights and roundabouts 
(as these are easy to access from \gls{osm}).
Modify \gls{gtfs} data
by adding \texttt{intersections}, \texttt{segments}, and \texttt{shape\_segments} tables.
We assume that if two routes both go through intersections $A$ and $B$,
they follow the same road to do so
(once we scale up to all intersections, this will be true almost surely).
We also allow segments to begin or end at a bus stop, 
which allows routes with the same origin or destination to share segments.

@@

In our particle filter implementation,
for a single vehicle $v$ we initialize $I$ particles,
and using the superscript $(i) : i = 1, \ldots, I$ to denote 
a value for a single particle.
Each particle is initialized with a full trajectory (see below),
from the begining of the route at time $t_o^{(i)}$ (trip start time is unknown),
until they reach the last stop $s_J$.

@@

In our chosen approach, a particle represents a full journey,
which allows us to estimate travel times for previous segments, 
as well as produce arrival time predictions in the same step.
The trajectory, which is demonstrated in figure~\ref{fig:trajectory},
is computed at each second of the journey, started at $t_0^{(i)}$.
Each particle is indendently run through a \emph{transition} algorithm,
outlined in figure~\ref{alg:transition}.

<<trajectory,echo=FALSE,out.width='\\linewidth',fig.cap="desc",fig.scap="short desc",fig.width=10,fig.height=5>>=
plot.new()
plot.window(c(0, 600), c(0, 1000), xaxs = "i", yaxs = "i")
axis(2, c(0, 400, 600, 1100), c(0, expression(s[1]^d), expression(r[1]^d), ""), las = 1, family = "serif")
axis(1, c(0, 700), c(expression(t[0]), ""))
mtext("Time (s)", 1, 2)
abline(h = 400, lty = 2, col = "#999999")
abline(h = 600, lty = "7727", lwd = 2, col = "#999999")

\vspace{1em}\hrule\vspace{1em}

@@

\begin{xtabular}{ll}
  \toprule
  &\textbf{Data} \\ \midrule
  $t_k$, $t_c$ & time at observation $k$, $c$ \\
  $\delta_x$ & time since last observation, $t_x - t_{x-1}$\\
  $\bY_k$ & observation $k$ \\
  $\tilde y_k$ & GPS position of vehicle (degrees) \\
  $\phi,\lambda$ & latitude, longitude (radians) \\
  $T^a_j, T^d_j$ & arrival, departure time at stop $j \in \{1, \ldots, J\}$ \\
  $S_n$ & unique identifier of stop $n$ \\
  $s_j$, $s_j^x$, $s_j^p$ & unique identifier of $j$th stop (of relevent route), shape distance traveled, position \\
  $R_m$ & unique identifier of segment $m$  \\
  $r_\ell$, $r_\ell^x$, $r_\ell^p$ & unique identifier of $\ell$th segment (of relevent route), shape distance traveled, path \\
  \bottomrule
  \\ \toprule
  &\textbf{Parameters} \\ \midrule
  $\pi$ & probability of stopping at a stop \\
  $\gamma$ & minimum dwell time (seconds) \\
  $\tau$ & mean dwell time (seconds) \\
  $\rho$ & probability of stopping at an intersection \\
  $\theta_m$ & mean queue time (seconds) at intersection of segment $m$ \\
  $\beta_m$ & mean travel time (seconds) for segment $m$ \\
  $\bbeta$, $\Sigma$ & transit network state (mean, covariance) \\
  $\eta$ & stop size (m) \\
  $\kappa$ & intersection size (m) \\
  $\sigma_x$ & system noise (variability in speed) \\
  $\sigma_y$ & GPS error \\
  $\varepsilon$ & arrival/departure time error \\
  $\xi$ & system noise (Kalman filter) \\
  \bottomrule
  \\ \toprule
  &\textbf{Variables} \\ \midrule
  $\bX_k$ & vehicle state associated with obs $k$ \\
  $x$, $\dot x$ & vehicle distance into trip, velocity \\
  $j$, $\ell$ & stop, segment index \\
  $A_j$, $\tilde D_j$, $D_j$ & arrival, dwell, departure time at stop $j$ \\
  $q_j$, $b_j$ & queue, travel time \\
  $z^{(i)}$ & computed GPS position of particle $(i)$ \\
  $p^{(i)}$ & indicator to whether particle $(i)$ stopped \\
  $w$ & particle weight \\
  $B_m$, $Q_m$ & mean, covariance of travel time of all recent vehicles along segment $m$ \\
  $\mat F_c$ & network state transition matrix \\
  $V$, $v$ & number of vehicles, indexed by $v$ \\
  $I$, $i$ & number of particles (per vehicle), indexed by $i$ \\
  $J$, $j$ & number of stops (conditional on route), indexed by $j$ \\
  $L$, $\ell$ & number of segments (conditional on route), indexed by $\ell$ \\
  $N$, $n$ & total number of stops in network, indexed by $n$ \\
  $M$, $m$ & total number of segments in network, indexed by $m$ \\
  $k$ & index for vehicle observations \\
  $c$ & index for transit network state updates \\
  \bottomrule
  \\ \toprule
  &\textbf{Functions} \\ \midrule
  $f$ & vehicle transition function \\
  $h$ & vehicle measurement function \\
  $g$ & equirectangular projection \\
  \bottomrule
\end{xtabular}

### 2017-08-03 realtime-filter/realtime-filter.tex

>>>>

In our chosen approach, a particle represents a full journey,
which allows us to estimate travel times for previous segments, 
as well as produce arrival time predictions in the same step.
The trajectory, which is demonstrated in figure~\ref{fig:trajectory},
is computed at each second of the journey, started at $t_0^{(i)}$.
Each particle is indendently run through a \emph{transition} algorithm,
outlined in figure~\ref{alg:transition}.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\linewidth]{figure/trajectory-1} \caption[short desc]{desc}\label{fig:trajectory}
\end{figure}

\vspace{1em}\hrule\vspace{1em}
